---
title: "Weighted log-rank tests"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{weighted_log_rank_tests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

See Magirr and Burman (2019) and Magirr (2021) for details about the weighted log-rank tests, and in particular the modestly weighted log-rank test. This vignette simply contains a summary of the formulae used within this package.

Firstly, the general definition of the weighted log-rank test is provided. Consider the ordered, distinct event times $t_1, \dots, t_k$. Let $d_{0,j}$ and $d_{1,j}$ be the number of events at event time $t_{j}$ on each of the arms respectively, and let $d_{j}$ be equal to the sum of these two values. Similarly, let $n_{0,j}$ and $n_{1,j}$ be the number at risk at event time $t_{j}$ on each of the arms respectively, and let $n_{j}$ be equal to the sum of these two values. At each time $t_{j}$, we evaluate the observed number of events on one arm, $d_{0,j}$, and the expected number of events on the same arm, $d_j \frac{n_{0,j}}{n_j}$. The test statistic $U$ is then a weighted sum of the difference of these values, i.e.

\[
U = \sum_{j=1}^k w_j \left(d_{0,j} - d_j \frac{n_{0,j}}{n_j}\right)
\]

The weights $w_j$ that are used depend on the type of weighted log-rank test. These are described in the next section.

Under the assumption that the survival curves of the two treatment arms are equal, the distribution of $U$ is

\[
U \sim N\left( 0, \sum_{j=1}^k w_j^2\frac{n_{0,j}n_{1,j} d_j (n_j - d_j)}{n_j^2(n_j-1)}\right)
\]


## Weights

- The standard log-rank test uses weights:
\[
w_j=1
\]

- The Fleming-Harrington ($\rho$,$\gamma$) test uses weights:
\[
w_j = \hat{S}(t_{j}-) ^ \rho * (1 - \hat{S}(t_{j}-)) ^ \gamma
\]
where $\hat{S}(t)$ is the Kaplan Meier estimate of the survival curve in the pooled data (both treatment arms) and time $t_j-$ is the time just before $t_j$. 
There is the matter of choosing $\rho$ and $\gamma$. A popular choice is $\rho=0$ and $\gamma=1$ which means that the weights are equal to 1 minus the Kaplan Meier estimate of the survival curve.

- The modestly weighted log-rank test uses weights:
\[
w_j = 1 / \max{\{\hat{S}(t_{j}-), \hat{S}(t^\ast)\}}
\]
There is the matter of choosing $t^\ast$ or alternatively choosing the value of $\hat{S}(t^\ast)$. See Magirr (2021) for a discussion on this.

## Permutation test and scores

Leton and Zuluaga (2001) showed that every weighted log-rank test can be written as either
an observed-minus-expected test (as described above), or as a permutation test.


The weights can be rewritten as scores for a permutation test using the following formulae for the censoring scores and event scores respectively:

\[
C_j=-\sum_{i=1}w_i\frac{d_i}{n_i}
\]

\[
c_j=C_j+w_j
\]


## Using the package


First, we simulate some data using the `sim_events_delay` function. 

This function simulates a dataset for a two-arm RCT with delayed separation of survival curves. The survival times on the control arm
follow an exponential distribution, and the survival times on the experimental arm follow a two-piece exponential distribution.
Given rate parameter $\lambda$, the exponential distribution has the form:

\[
f(t)=\lambda \exp(-\lambda t)
\]

Recruitment onto the trial is modeled using the power model
\[
P(\mathrm{recruited\ before\ T}) = (\mathrm{T} / \mathrm{rec\_period}) ^ \mathrm{rec\_power}
\]
where
$\mathrm{rec\_period}$ is the time at the end of recruitment period, and $\mathrm{rec\_power}$ controls the rate of recruitment.

This means that participants enter the trial at calendar times after time 0. However, all observations are censored at the calendar time `max_cal_t`.


```{r 1}
library(wlrt)
set.seed(1)
sim_data <- sim_events_delay(
  n_c = 5,
  n_e = 5,
  delay_e = 6,
  lambda_c = log(2)/9,
  lambda_e_1 = log(2)/9,
  lambda_e_2 = log(2)/18,
  rec_period = 12,
  rec_power = 1,
  max_cal_t = 36
)
sim_data
```


```{r find_scores}
scores_out<-find_scores(data=sim_data,
  formula=Surv(event_time,event_status)~group,
  wlr="mw",
  t_star = 3
)
plot(scores_out)

```

## References
Leton, E. and Zuluaga, P. (2001) 
Equivalence between score and weighted tests for survival curves. 
Commun Stat., 30(4), 591-608.

Magirr, D. (2021).
Non-proportional hazards in immuno-oncology: Is an old perspective needed?.
Pharmaceutical Statistics, 20(3), 512-527.

Magirr, D. and Burman, C.F., (2019).
Modestly weighted logrank tests.
Statistics in medicine, 38(20), 3782-3790.
